---
title: "Parsing Chess Tournament Results into a Tidy CSV"
author: "Esra Dogan"
date: "2026-09-22"
output: html_document
---

# Introduction

In this project, I work with a text file containing the results of a chess tournament with 64 players. The file contains each player's pair number, name, state, total points, USCF ID, pre and post tournament ratings, and the result and opponent of each of the seven rounds. However, the file is designed to be read by people, not by software: each player's information is split across two lines, fields are separated by pipe characters (\|) and padded with spaces and dashed lines separate the records.

In other words, the data is not tidy. According to Wickham, Çetinkaya-Rundel, and Grolemund (2023), a dataset is tidy when each variable is a column, each observation is a row, and each value is a cell. In this file, a single observation (one player) spans two lines, and some cells hold more than one value. For example, in player 8's record (Ezekiel Houghton), the cell "W 3" holds both the result and the opponent's pair number, and "R: 1641P17-\>1657P24" holds the pre-rating, the post-rating, and provisional rating markers all in one place.

The goal is to transform this file into a tidy CSV file with one row per player and five columns: player's name, state, total points, pre-tournament rating, and average pre-tournament rating of opponents. The last column does not exist in the file and must be calculated.

The chess rating system developed by Arpad Elo has been used in many other contexts, including evaluating job candidates. A player's total points alone do not tell the full story, because scoring 5 points against strong opponents is more impressive than scoring 5 points against weak ones. The average opponent rating adds this context. As an extension, I will compare each player's own rating with the average rating of their opponents.

## Planned Approach

I will read the text file into R from GitHub and reshape it so that each player's information is in a single row. Then I will extract the player's name, state, total points, and pre-rating, along with the opponents they played.

To calculate the average opponent rating, I will look up each opponent's pre-rating and average it over the games actually played. Finally, I will check the results against the assignment example, export the table as a CSV file, and compare each player's rating with the average rating of their opponents.

## Anticipated Challenges 
The main challenge is calculating the average opponent rating correctly. Byes and unplayed rounds have no opponent, so they must be left out, and players who played fewer than seven games should be averaged only over the games they actually played. Since each opponent's rating is stored in that opponent's own row, all ratings need to be extracted before any averages can be calculated.

Cleaning the file also has its own difficulties. Each player's information is split across two lines that must be matched correctly, and some cells hold more than one value as I mentioned before.


```{r}
library(tidyverse)
library(gganimate)
```

```{r}
chess_raw <- read_lines("https://raw.githubusercontent.com/esradogan3/data607-project1/refs/heads/main/tournamentinfo.txt")
```

```{r}
head(chess_raw)
length(chess_raw)
```
```{r}
url <- "https://raw.githubusercontent.com/esradogan3/data607-project1/refs/heads/main/tournamentinfo.txt"

chess_raw <- readLines(url, warn = FALSE) # hide the warning while reading lines
chess_raw <- chess_raw[!grepl("^-+\\s*$", chess_raw)] # drop the dashed separator lines
chess_raw <- chess_raw[-(1:2)] # drop the two header lines (Pair/Num rows)

row1 <- chess_raw[seq(1, length(chess_raw), by = 2)] # pair num, name, points, rounds
row2 <- chess_raw[seq(2, length(chess_raw), by = 2)] # state, USCF ID / ratings

p1 <- read.table(text = row1, sep = "|", strip.white = TRUE,
                 quote = "", stringsAsFactors = FALSE)
p2 <- read.table(text = row2, sep = "|", strip.white = TRUE,
                 quote = "", stringsAsFactors = FALSE)

chess <- data.frame(
  id         = p1$V1,
  name       = p1$V2,
  state      = p2$V1,
  points     = as.numeric(p1$V3),
  pre_rating = as.numeric(sub(".*R:\\s*(\\d+).*", "\\1", p2$V2))
)

print(chess)
```
**Checking if everything looks good:**
```{r}
nrow(chess)
View(chess)
```
## Step by Step Calculating Average Pre-Ratings of Opponent's:
```{r}
gsub("\\D", "", unlist(p1[1, 4:10]))
```
```{r}
opponents <- as.integer(gsub("\\D", "", unlist(p1[1, 4:10])))
opponents
```
```{r}
opp_ratings <- chess$pre_rating[opponents]
opp_ratings
mean(opp_ratings, na.rm = TRUE)
```
```{r}
opp_avg <- numeric(64)

for (i in 1:64) {
  opponents   <- as.integer(gsub("\\D", "", unlist(p1[i, 4:10])))
  opp_ratings <- chess$pre_rating[opponents]
  opp_avg[i]  <- mean(opp_ratings, na.rm = TRUE)
}
```

```{r}
chess <- mutate(chess, opp_avg_prerating = opp_avg)

head(chess)
```
```{r}
print(chess[c("pre_rating", "opp_avg_prerating")])
```
The table above places each player's pre-rating next to their opponents' average pre-rating, so the two can be compared directly. I discuss this comparison in more detail in the video.

```{r}
write.csv(chess, "chess_ratings.csv", row.names = FALSE)
```
## Additional Exploration: How the Standings Changed Round by Round

Beyond the main task, I wanted to see how the tournament unfolded over time. Using each player's result in every round, I calculated their running point total after each of the 7 rounds and animated the top 10 players' standings with gganimate. As a check, each player's total after round 7 matches their final `points` value in the data.

```{r}
# Step 1: Get each round's result letter (W, L, D, ...) from columns V4 to V10
rounds <- p1[, c("V4", "V5", "V6", "V7", "V8", "V9", "V10")]

for (i in 1:7) {
  rounds[[i]] <- substr(trimws(rounds[[i]]), 1, 1)
}

rounds$id     <- 1:nrow(rounds)
rounds$player <- chess$name

# Step 2: Convert letters to points (win = 1, draw = 0.5, loss = 0)
points_df <- rounds

for (i in 1:7) {
  letter <- points_df[[i]]
  points_df[[i]] <- ifelse(letter %in% c("W", "X", "B"), 1,
                    ifelse(letter %in% c("D", "H"), 0.5, 0))
}

# Step 3: Running total of points after each round
cum_df <- points_df

for (i in 2:7) {
  cum_df[[i]] <- cum_df[[i - 1]] + points_df[[i]]
}

# Step 4: Reshape to long format (one row per player per round)
long <- pivot_longer(cum_df,
                     cols = V4:V10,
                     names_to = "round",
                     values_to = "total")

long$round <- as.numeric(substr(long$round, 2, 3)) - 3

# Step 5: Keep the top 10 players in each round
top <- NULL

for (r in 1:7) {
  this_round <- long[long$round == r, ]
  this_round <- this_round[order(-this_round$total), ]
  this_round$rank <- 1:nrow(this_round)
  top <- rbind(top, this_round[1:10, ])
}

# Step 6: Animated bar chart of the standings, round by round
p <- ggplot(top, aes(x = rank, y = total, group = player, fill = player)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = player), hjust = -0.05, size = 3) +
  coord_flip(clip = "off") +
  scale_x_reverse() +
  scale_y_continuous(limits = c(0, 9)) +
  labs(title = "Standings after round {closest_state}",
       x = NULL, y = "Total points") +
  transition_states(round, transition_length = 2, state_length = 1)

animate(p, nframes = 100, fps = 10)
```
```
## Using LLM:
I used Claude (Anthropic, 2026) as a tutor while working on this project, to understand R functions and regular expressions, debug errors, and plan the calculation and animation steps. I wrote the code myself with its guidance. The link to the chat: https://claude.ai/share/ef57a130-5f58-4300-a86c-ca45e3cfdd47

## References

Anthropic. (2026). Claude (Opus 5.5 version) [Large language model]. https://claude.ai
Wickham, H., Çetinkaya-Rundel, M., & Grolemund, G. (2023). *R for Data Science* (2nd ed.). O'Reilly Media. https://r4ds.hadley.nz